Creating an Arrow IPC file from the NYC 311 data

### 2023-06-17 nyc311clean.qmd
Even more tweaking.

---
title: "Creating an Arrow IPC file from the NYC 311 data"
author: 
  - name: Jun Yan
    affiliation: University of Connecticut
  - name: Douglas Bates
    affiliation: University of Wisconsin - Madison
---

@@

::: {.callout-note collapse="false"}
### Need to scroll right to see the entire line

@@

and, of course, the problem is that someone got creative and used `na` as missing value indicator.

@@

These rows can be isolated for manual checking

@@

```{julia}
first(df[earlyclosing, :], 3)
```

@@

@@

This is a tricky question because the values of `Latitude` and `Longitude` are stored to 17 significant digits, which is much more accuracy than could possibly be measured.
To check for consistency we need to convert the `Location` to a `Latitude-Longitude` pair then compare these values to the recorded `Latitude` and `Longitude` with some tolerance.

@@

```{julia}
#| output: false
fn = Arrow.write("./data/311_half_March_2023.arrow", df)
```

The Arrow file is less than half the size of the original CSV file but contains much more specificity about the data types and representations.
Furthermore it can be read into other data science environments without needing to tinker with optional arguments.

The Arrow IPC format and the `Arrow.write` function in the Julia package allow for optional compression of the contents using `lz4` or `zstd` if further reduction of the file size is desired.
The trade-off is that the uncompressed file can be memory-mapped by some systems to make for extremely fast input.

@@

or R
```r
> tibble::glimpse(arrow::read_ipc_file("./data/311_half_March_2023.arrow"))
Rows: 125,826
Columns: 46
$ Unique_Key                     <int> 57000688, 56993837, 57049479, 57056381,…
$ Created_Date                   <dttm> 2023-03-09 09:05:25, 2023-03-08 15:27:…
$ Closed_Date                    <dttm> 2023-03-13 03:21:11, 2023-05-03 02:17:…
$ Agency                         <fct> DOE, TLC, HPD, NYPD, DOB, DPR, DOT, NYP…
$ Agency_Name                    <fct> Department of Education, Taxi and Limou…
$ Complaint_Type                 <fct> School Maintenance, For Hire Vehicle Co…
$ Descriptor                     <chr> "Other School Condition", "Driver Compl…
$ Location_Type                  <fct> School, Street, RESIDENTIAL BUILDING, S…
$ Incident_Zip                   <int> NA, 10075, 10459, 10455, 11365, 11691, …
$ Incident_Address               <chr> NA, "995 MADISON AVENUE", "1010 BRYANT …
$ Street_Name                    <chr> NA, "MADISON AVENUE", "BRYANT AVENUE", …
$ Cross_Street_1                 <chr> NA, "EAST   77 STREET", NA, "EAST  147 …
$ Cross_Street_2                 <chr> NA, "EAST   78 STREET", NA, "EAST  149 …
$ Intersection_Street_1          <chr> NA, "EAST   77 STREET", NA, "EAST  147 …
$ Intersection_Street_2          <chr> NA, "EAST   78 STREET", NA, "EAST  149 …
$ Address_Type                   <fct> NA, ADDRESS, ADDRESS, ADDRESS, ADDRESS,…
$ City                           <fct> NA, NEW YORK, BRONX, BRONX, FRESH MEADO…
$ Landmark                       <chr> NA, "MADISON AVENUE", NA, "UNION AVENUE…
$ Facility_Type                  <fct> NA, NA, NA, NA, NA, NA, N/A, NA, NA, NA…
$ Status                         <fct> Closed, Closed, Closed, Closed, Assigne…
$ Due_Date                       <dttm> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA…
$ Resolution_Description         <fct> "The Department of Education determined…
$ Resolution_Action_Updated_Date <dttm> 2023-03-13 03:21:17, 2023-05-03 02:17:…
$ Community_Board                <fct> Unspecified BROOKLYN, 08 MANHATTAN, 02 …
$ BBL                            <int64> NA, 1013927501, 2027560001, 202582004…
$ Borough                        <fct> BROOKLYN, MANHATTAN, BRONX, BRONX, QUEE…
$ X_Coordinate_State_Plane_      <int> 989013, 994507, 1015090, 1010420, 10388…
$ Y_Coordinate_State_Plane_      <int> 173052, 221687, 239359, 235260, 206195,…
$ Open_Data_Channel_Type         <fct> PHONE, ONLINE, PHONE, MOBILE, UNKNOWN, …
$ Park_Facility_Name             <fct> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,…
$ Park_Borough                   <fct> BROOKLYN, MANHATTAN, BRONX, BRONX, QUEE…
$ Vehicle_Type                   <fct> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,…
$ Taxi_Company_Borough           <fct> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,…
$ Taxi_Pick_Up_Location          <chr> NA, "995 MADISON AVENUE, MANHATTAN (NEW…
$ Bridge_Highway_Name            <fct> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,…
$ Bridge_Highway_Direction       <fct> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,…
$ Road_Ramp                      <fct> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,…
$ Bridge_Highway_Segment         <fct> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,…
$ Latitude                       <dbl> 40.64166, 40.77515, 40.82361, 40.81237,…
$ Longitude                      <dbl> -73.98284, -73.96297, -73.88857, -73.90…
$ Location                       <chr> "(40.641663101476226, -73.9828374867858…
$ Zip_Codes                      <int> 17620, 10092, 10937, 10933, 14507, 2052…
$ Community_Districts            <int> 2, 23, 8, 49, 25, 51, NA, 26, 5, 24, 47…
$ Borough_Boundaries             <int> 2, 4, 5, 5, 3, 3, NA, 3, 2, 5, 4, 5, 3,…
$ City_Council_Districts         <int> 27, 51, 43, 35, 24, 47, NA, 16, 8, 29, …
$ Police_Precincts               <int> 39, 11, 24, 23, 65, 59, NA, 69, 38, 34,…
:::{.callout-note collapse="true"}
### R applies a time zone to the datetime values
The R `arrow` package appears to read the DateTime values as if they are in given in UTC then prints them in the local time zone.
:::

or Python (not that this format is called `feather` in the `pyarrow` package)

@@

>>> fea.read_table("./data/311_half_March_2023.arrow")

@@

Location: string

@@

>>>

@@

>>>>

---
title: "Creating an Arrow IPC file from the NYC 311 data"
author: 
  - name: Jun Yan
    affiliation: University of Connecticut
  - name: Douglas Bates
    affiliation: University of Wisconsin - Madison
jupyter: julia-1.9
execute:
  cache: true
  freeze: auto
---

@@

::: {.callout-note collapse="false"}
### Need to scroll right to see all of the output lines

@@

and, of course, the problem is that someone got creative and used `na` as a missing value indicator.

@@

We save the indicators of rows with early closing or immediate closing for later manual checking.

@@

```{julia}
earlyfrm = df[earlyclosing, :]
first(earlyfrm, 3)
```

@@

These three are from the "Department of Transport" and, in fact, all rows with early closing dates are from the DOT

```{julia}
unique(select(earlyfrm, :Agency, :Agency_Name))
```

@@

This is a tricky question because the values of `Latitude` and `Longitude` are stored to 17 significant digits, which is much more accuracy than could possibly be measured.
To check for consistency we need to convert the `Location` to a `Latitude-Longitude` pair then compare these values to the recorded `Latitude` and `Longitude` by employing a tolerance in the comparison of floating-point values.
We'll leave that as an exercise for the reader and just assure you that the `Location` column is redundant and can be dropped.

@@

```{julia}
#| output: false
fn = Arrow.write(
    "./data/311_half_Mar_2023.arrow",
    select(df, Not(:Location)),
)
```

The Arrow file is less than one third the size of the original CSV file but contains much more specificity about the data types and representations.
Furthermore it can be read into other data science environments without needing to tinker with arguments to CSV file-reading functions.

The Arrow IPC format and the `Arrow.write` function in the Julia package allow for optional compression of the contents using `lz4` or `zstd` compression, if further reduction of the file size is desired.
The downside of compression is that an uncompressed file can be memory-mapped, allowing for extremely fast input, whereas the compressed files must be decompressed into memory.

@@

or R
```r
> tibble::glimpse(arrow::read_ipc_file("./data/311_half_Mar_2023.arrow"))
Rows: 125,826
Columns: 45
$ Unique_Key                     <int> 57000688, 56993837, 57049479, 57056381,…
$ Created_Date                   <dttm> 2023-03-09 09:05:25, 2023-03-08 15:27:…
$ Closed_Date                    <dttm> 2023-03-13 03:21:11, 2023-05-03 02:17:…
$ Agency                         <fct> DOE, TLC, HPD, NYPD, DOB, DPR, DOT, NYP…
$ Agency_Name                    <fct> Department of Education, Taxi and Limou…
$ Complaint_Type                 <fct> School Maintenance, For Hire Vehicle Co…
$ Descriptor                     <chr> "Other School Condition", "Driver Compl…
$ Location_Type                  <fct> School, Street, RESIDENTIAL BUILDING, S…
$ Incident_Zip                   <int> NA, 10075, 10459, 10455, 11365, 11691, …
$ Incident_Address               <chr> NA, "995 MADISON AVENUE", "1010 BRYANT …
$ Street_Name                    <chr> NA, "MADISON AVENUE", "BRYANT AVENUE", …
$ Cross_Street_1                 <chr> NA, "EAST   77 STREET", NA, "EAST  147 …
$ Cross_Street_2                 <chr> NA, "EAST   78 STREET", NA, "EAST  149 …
$ Intersection_Street_1          <chr> NA, "EAST   77 STREET", NA, "EAST  147 …
$ Intersection_Street_2          <chr> NA, "EAST   78 STREET", NA, "EAST  149 …
$ Address_Type                   <fct> NA, ADDRESS, ADDRESS, ADDRESS, ADDRESS,…
$ City                           <fct> NA, NEW YORK, BRONX, BRONX, FRESH MEADO…
$ Landmark                       <chr> NA, "MADISON AVENUE", NA, "UNION AVENUE…
$ Facility_Type                  <fct> NA, NA, NA, NA, NA, NA, N/A, NA, NA, NA…
$ Status                         <fct> Closed, Closed, Closed, Closed, Assigne…
$ Due_Date                       <dttm> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA…
$ Resolution_Description         <fct> "The Department of Education determined…
$ Resolution_Action_Updated_Date <dttm> 2023-03-13 03:21:17, 2023-05-03 02:17:…
$ Community_Board                <fct> Unspecified BROOKLYN, 08 MANHATTAN, 02 …
$ BBL                            <int64> NA, 1013927501, 2027560001, 202582004…
$ Borough                        <fct> BROOKLYN, MANHATTAN, BRONX, BRONX, QUEE…
$ X_Coordinate_State_Plane_      <int> 989013, 994507, 1015090, 1010420, 10388…
$ Y_Coordinate_State_Plane_      <int> 173052, 221687, 239359, 235260, 206195,…
$ Open_Data_Channel_Type         <fct> PHONE, ONLINE, PHONE, MOBILE, UNKNOWN, …
$ Park_Facility_Name             <fct> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,…
$ Park_Borough                   <fct> BROOKLYN, MANHATTAN, BRONX, BRONX, QUEE…
$ Vehicle_Type                   <fct> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,…
$ Taxi_Company_Borough           <fct> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,…
$ Taxi_Pick_Up_Location          <chr> NA, "995 MADISON AVENUE, MANHATTAN (NEW…
$ Bridge_Highway_Name            <fct> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,…
$ Bridge_Highway_Direction       <fct> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,…
$ Road_Ramp                      <fct> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,…
$ Bridge_Highway_Segment         <fct> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,…
$ Latitude                       <dbl> 40.64166, 40.77515, 40.82361, 40.81237,…
$ Longitude                      <dbl> -73.98284, -73.96297, -73.88857, -73.90…
$ Zip_Codes                      <int> 17620, 10092, 10937, 10933, 14507, 2052…
$ Community_Districts            <int> 2, 23, 8, 49, 25, 51, NA, 26, 5, 24, 47…
$ Borough_Boundaries             <int> 2, 4, 5, 5, 3, 3, NA, 3, 2, 5, 4, 5, 3,…
$ City_Council_Districts         <int> 27, 51, 43, 35, 24, 47, NA, 16, 8, 29, …
$ Police_Precincts               <int> 39, 11, 24, 23, 65, 59, NA, 69, 38, 34,…
```
:::{.callout-note collapse="true"}
### R applies a time zone to the datetime values
The R `arrow` package appears to read the DateTime values as if they are UTC values then prints them in the local time zone.
:::

or in Python with `pyarrow` or with [polars](https://pola.rs).
(Note that the Arrow IPC format is called `feather` in the `pyarrow` package.)

@@

>>> import polars as pl
>>> fea.read_table('./data/311_half_Mar_2023.arrow')

@@

@@

>>> pl.read_ipc('./data/311_half_Mar_2023.arrow', use_pyarrow=True)
shape: (125_826, 45)
┌──────────┬────────────┬───────────┬────────┬───┬────────────┬────────────┬────────────┬────────────┐
│ Unique_K ┆ Created_Da ┆ Closed_Da ┆ Agency ┆ … ┆ Community_ ┆ Borough_Bo ┆ City_Counc ┆ Police_Pre │
│ ey       ┆ te         ┆ te        ┆ ---    ┆   ┆ Districts  ┆ undaries   ┆ il_Distric ┆ cincts     │
│ ---      ┆ ---        ┆ ---       ┆ cat    ┆   ┆ ---        ┆ ---        ┆ ts         ┆ ---        │
│ i32      ┆ datetime[m ┆ datetime[ ┆        ┆   ┆ i8         ┆ i8         ┆ ---        ┆ i8         │
│          ┆ s]         ┆ ms]       ┆        ┆   ┆            ┆            ┆ i8         ┆            │
╞══════════╪════════════╪═══════════╪════════╪═══╪════════════╪════════════╪════════════╪════════════╡
│ 57000688 ┆ 2023-03-09 ┆ 2023-03-1 ┆ DOE    ┆ … ┆ 2          ┆ 2          ┆ 27         ┆ 39         │
│          ┆ 15:05:25   ┆ 3         ┆        ┆   ┆            ┆            ┆            ┆            │
│          ┆            ┆ 08:21:11  ┆        ┆   ┆            ┆            ┆            ┆            │
│ 56993837 ┆ 2023-03-08 ┆ 2023-05-0 ┆ TLC    ┆ … ┆ 23         ┆ 4          ┆ 51         ┆ 11         │
│          ┆ 21:27:34   ┆ 3         ┆        ┆   ┆            ┆            ┆            ┆            │
│          ┆            ┆ 07:17:20  ┆        ┆   ┆            ┆            ┆            ┆            │
│ 57049479 ┆ 2023-03-15 ┆ 2023-03-2 ┆ HPD    ┆ … ┆ 8          ┆ 5          ┆ 43         ┆ 24         │
│          ┆ 19:26:41   ┆ 1         ┆        ┆   ┆            ┆            ┆            ┆            │
│          ┆            ┆ 21:16:53  ┆        ┆   ┆            ┆            ┆            ┆            │
│ 57056381 ┆ 2023-03-16 ┆ 2023-03-1 ┆ NYPD   ┆ … ┆ 49         ┆ 5          ┆ 35         ┆ 23         │
│          ┆ 23:58:53   ┆ 7         ┆        ┆   ┆            ┆            ┆            ┆            │
│          ┆            ┆ 02:44:31  ┆        ┆   ┆            ┆            ┆            ┆            │
│ …        ┆ …          ┆ …         ┆ …      ┆ … ┆ …          ┆ …          ┆ …          ┆ …          │
│ 56954525 ┆ 2023-03-04 ┆ 2023-03-0 ┆ DEP    ┆ … ┆ 12         ┆ 4          ┆ 10         ┆ 6          │
│          ┆ 16:23:00   ┆ 8         ┆        ┆   ┆            ┆            ┆            ┆            │
│          ┆            ┆ 10:50:00  ┆        ┆   ┆            ┆            ┆            ┆            │
│ 57132547 ┆ 2023-03-24 ┆ 2023-03-2 ┆ NYPD   ┆ … ┆ 30         ┆ 1          ┆ 14         ┆ 76         │
│          ┆ 21:57:44   ┆ 4         ┆        ┆   ┆            ┆            ┆            ┆            │
│          ┆            ┆ 23:18:11  ┆        ┆   ┆            ┆            ┆            ┆            │
│ 57102740 ┆ 2023-03-21 ┆ 2023-03-2 ┆ NYPD   ┆ … ┆ 70         ┆ 4          ┆ 32         ┆ 4          │
│          ┆ 05:10:12   ┆ 1         ┆        ┆   ┆            ┆            ┆            ┆            │
│          ┆            ┆ 05:16:30  ┆        ┆   ┆            ┆            ┆            ┆            │
│ 56950149 ┆ 2023-03-04 ┆ 2023-03-0 ┆ DOT    ┆ … ┆ 41         ┆ 3          ┆ 16         ┆ 61         │
│          ┆ 00:27:00   ┆ 4         ┆        ┆   ┆            ┆            ┆            ┆            │
│          ┆            ┆ 10:15:00  ┆        ┆   ┆            ┆            ┆            ┆            │
└──────────┴────────────┴───────────┴────────┴───┴────────────┴────────────┴────────────┴────────────┘

@@

The columns that are described as `<fct>` in the `glimpse` output from R are characterized as `dictionary` in the output from `fea.read_table` and listed as `cat` (for "categorical") in the Polars data frame.
The conversion of a column of character strings, many of which are repetitions, to a combination of a vector of the unique values and a small-sized integer index into that vector, can result in considerable storage saving, at the expense of a trivial amount of processing.
For example, storing the `Agency_Name` as a vector of strings takes about 34 MB but, after dictionary encoding, it takes up about 1/8 MB.
